---
title: "Practical A"
author: "Gerko Vink"
date: ""
output: 
   html_document:
    toc: true
    toc_depth: 5
    toc_float: true
    number_sections: false
---
  
<style type="text/css">
  
body{ /* Normal  */
  font-size: 12px;
  }
td {  /* Table  */
  font-size: 12px;
}
h1.title {
  font-size: 18px;
  color: DarkBlue;
}
h1 { /* Header 1 */
  font-size: 18px;
}
h2 { /* Header 2 */
  font-size: 18px;
}
h3 { /* Header 3 */
  font-size: 18px;
}
code.r{ /* Code block */
  font-size: 12px;
}
pre { /* Code block - determines code spacing between lines */
  font-size: 14px;
}
</style>
  
---
  
```{r echo=FALSE}
printAnswers <- TRUE
```

We use the following packages in this Practical:
```{r message=FALSE, warning = FALSE}
library(dplyr)     # for data manipulation
library(magrittr)  # for pipes
library(ggplot2)   # for visualization
library(mice)      # for the boys data
```

---

# Exercises

---

## Pipes

--- 

1. **Use a pipe to do the following:**

- draw 1000 values from a normal distribution with `mean = 5` and `sd = 1` - $N(5, 1)$, 
- create a matrix where the first 500 values are the first column and the second 500 values are the second column **
- make a scatterplot of these two columns

---

2. **Use a pipe to calculate the correlation matrix on the `anscombe` data set**

---

3. **Now use a pipe to calculate the correlation for the pair (`x4`, `y4`) on the `anscombe` data set**

---

The `boys` dataset is part of package `mice`. It is a subset of 748 Dutch boystaken from the Fourth Dutch Growth Study. It's columns record a variety of growth measures. Inspect the help for `boys` dataset and make yourself familiar with its contents.**

To learn more about the contents of the data, use one of the two following help commands:
```{r}
help(boys)
?boys
```

---

4. **It seems that the `boys` data are sorted based on `age`. Verify this.** 


---

5. **Use a pipe to calculate the correlation between `hgt` and `wgt` in the `boys` data set from package `mice`.**

---

6. **In the `boys` data set, `hgt` is recorded in centimeters. Use a pipe to transform `hgt` in the `boys` dataset to height in meters and verify the transformation**

---

7. **Use a pipe to plot the pair (`hgt`, `wgt`) two times: once for `hgt` in meters and once for `hgt` in centimeters. Make the points in the 'centimeter' plot `red` and in the 'meter' plot `blue`. **

---

## Visualization

---

8. **Function `plot()` is the core plotting function in `R`. Find out more about `plot()`: Try both the help in the help-pane and `?plot` in the console. Look at the examples by running `example(plot)`.**

---

9. **Create a scatterplot between `age` and `bmi` in the `mice::boys` data set**

---

10. **Now recreate the plot with the following specifications:**

---

11. **Create a histogram for `age` in the `boys` data set**

---

12. **Create a bar chart for `reg` in the boys data set**

---

13. **Create a box plot for `hgt` with different boxes for `reg` in the `boys` data set**

---

14. **Create a density plot for `age` with different curves for boys from the `city` and boys from rural areas (`!city`).**

---

15. **Create a diverging bar chart for `hgt` in the `boys` data set, that displays for every `age` year that year's mean height in deviations from the overall average `hgt`**

In  other words; recreate the following plot:

```{r echo=FALSE}
boys %>%
  mutate(Hgt = hgt - mean(hgt, na.rm = TRUE),
         Age = cut(age, 0:22, labels = 0:21)) %>%
  aggregate(Hgt ~ Age, data = ., mean) %>% #specify data = . to allow formula
  mutate(Diff = cut(Hgt, c(-Inf, 0, Inf),
                    labels = c("Below Average", "Above Average"))) %>%
  ggplot(aes(x = Age, y = Hgt, fill = Diff)) + 
  geom_bar(stat = "identity") +
  coord_flip()
```


---

End of Practical

---

# Useful References

- [The `ggplot2` reference page](https://ggplot2.tidyverse.org/reference/index.html)
- [`magrittr`](https://magrittr.tidyverse.org)
- [`R` for Data Science](http://r4ds.had.co.nz) - [Chapter 18 on pipes](http://r4ds.had.co.nz/pipes.html)
- Anscombe, Francis J. (1973) [Graphs in statistical analysis](http://www.sjsu.edu/faculty/gerstman/StatPrimer/anscombe1973.pdf). American Statistician, 27, 17–21.