---
title: "Exercise B"
author: "Gerko Vink"
date: "Rijkstrainees - Statistical Programming in R"
output: 
   html_document:
    toc: true
    toc_depth: 5
    toc_float: true
    number_sections: false
---
  
<style type="text/css">
  
body{ /* Normal  */
  font-size: 12px;
  }
td {  /* Table  */
  font-size: 12px;
}
h1.title {
  font-size: 18px;
  color: DarkBlue;
}
h1 { /* Header 1 */
  font-size: 18px;
}
h2 { /* Header 2 */
  font-size: 18px;
}
h3 { /* Header 3 */
  font-size: 18px;
}
code.r{ /* Code block */
  font-size: 12px;
}
pre { /* Code block - determines code spacing between lines */
  font-size: 14px;
}
</style>
  
---

In this practical I detail multiple skills and show you a workflow for (predictive) analytics. 

All the best, 

Gerko

---

# Exercises

---

The following packages are required for this practical:
```{r, message=FALSE}
library(dplyr)
library(magrittr)
library(mice)
library(ggplot2)
library(DAAG)
library(MASS)
```

---

## Exercise 1

---

The data sets `elastic1` and `elastic2` from the package `DAAG` were obtained
using the same apparatus, including the same rubber band, as the data frame
`elasticband`. 

1. **Using a different symbol and/or a different color, plot the data
from the two data frames `elastic1` and `elastic2` on the same graph. Do the two
sets of results appear consistent?**

---

## Exercise 2

---

2. **For each of the data sets `elastic1` and `elastic2`, determine the regression of
distance on stretch. In each case determine:**

- fitted values and standard errors of fitted values and
- the $R^2$ statistic.

Compare the two sets of results. What is the key difference between the two sets
of data?

---

## Exercise 3

---

3. **Study the *residual vs leverage* plots for both models. Hint use `plot()` on the fitted object**

---

Because there is a single value that influences the estimation and is somewhat different than the other values, a robust form of regression may be advisable to obtain more stable estimates. When robust methods are used, we refrain from omitting a suspected outlier from our analysis. In general, with robust analysis, influential cases that are not conform the other cases receive less weight in the estimation procedure then under non-robust analysis.

---

## Exercise 4

---

4. **Use the robust regression function `rlm()` from the `MASS` package to fit lines to the data in `elastic1` and `elastic2`. Compare the results with those from use of `lm()`:**

- residuals
- regression coefficients, 
- standard errors of coefficients, 
- plots of residuals against fitted values.

---

## Exercise 5

---

5. **Use the `elastic2` variable `stretch` to obtain predictions on the model fitted on `elastic1`.**

---

## Exercise 6

---

6. **Now make a scatterplot to investigate similarity between plot the predicted values against the observed values for `elastic2`**

---

The mammalsleep dataset is part of `mice`. It contains the Allison and Cicchetti (1976) data for mammalian species. To learn more about this data, type
```{r}
?mammalsleep
```

--- 

## Exercise 7

---

7. **Fit and inspect a model where `brw` is modeled from `bw`**

---

## Exercise 8

---

8. **Now fit and inspect a model where `brw` is predicted from both `bw` and `species`**

---

## Exercise 9

---

9. **Can you find a model that improves the $R^2$ in modeling `brw`?**

---

## Exercise 10

---

10. **Inspect the diagnostic plots for the model obtained in `exercise 16`. What issues can you detect?**

---

End of `Practical`. 
